Normalizations are not what you think; Explicit Scale Simulation in ALDEx2

### 2023-10-21 go3-author.tex
submitted version

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Beyond compositionally in high throughput sequencing; estimating
the importance of scale in data analysis with ALDEx2}
\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin Silverman$^{2,3}$}
\date{\today}

@@

$^1$ Department of Biochemistry, University of Western Ontario
$^2$ College of Information Sciences and Technology, Pennsylvania  State University
$^3$ Department of Medicine, Pennsylvania  State University

\begin{abstract}
Despite scale being an important confounding variable, high throughput
sequencing generates data that are difficult to reconcile with the scale
of the underlying environment, but it has been shown that the scale of
the environment is related to the geometric mean of the parts of the
sample. The ALDEx2 R package builds a Bayesian model of the data and
here we report an update to the model that includes both random sampling
and scale estimates. We show that incorporating scale allows for robust
inference in a number of different experimental designs and we provide
guidance on how to incorporate scale inference into differential
abundance analysis. While all data normalizations in widespread use make
implicit assumptions about the scale of the system we suggest that
explicit rather than implicit scale models should be used.
\end{abstract}

@@

High throughput sequencing (HTS) is a universal tool to explore many
biological phenomenon such as gene expression (single-cell sequencing,
RNA-sequencing, meta-transcriptomics), microbial community composition
(16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme
activity (selex, CRISPR killing). HTS proceeds by taking a fixed-number
sample from the environment, making a library, multiplexing (merging)
multiple libraries and applying a fixed number of molecules to the flow
cell. In essence it is a poll of the environment that is mixed with
other polls and then a poll of the mixture is taken. It should be clear
that the total number of molecules sequenced is driven by the capacity
of the instrument and not by the number of molecules in the sampled
environment(Lovell et al. 2015; Vandeputte et al. 2017; Props et al.
2017) .

Data that are generated by sequencing come from systems where scale is
usually important and may be a confounding variable (Lovell et al.
2015). For example, cells transformed by the cMyc oncogene have about 3
times the amount of mRNA and about twice the rRNA content than do
non-transformed cells (Nie et al. 2012), and this dramatically skews
transcriptome analysis (Lovén et al. 2012). In addition, wild-type and
mutant strains of cell lines, yeast or bacteria often have different
growth rates, which would affect our ability to identify truly
differentially abundant genes (Yoshikawa et al. 2011). As another
example, the total bacterial load of the vaginal microbiome differs by
1-2 orders of magnitude in absolute abundance between the healthy and
bacterial vaginosis states (Zozaya-Hinchliffe et al. 2010); the
composition is dramatically different as well (Ravel et al. 2011;
Hummelen et al. 2010). Thus, the full description of these systems
includes both relative change (composition) and absolute abundance
(scale) but current methods access only the compositional information.

More formally, if the true information in the environment being sampled
can be summarized by counts of features (genes, taxa, etc) in a matrix
\(\mathbf{W}\), then the elements of the matrix can be decomposed into
its composition (relative information, \(\mathbf{W}^{\parallel}\) ) and
its scale (total sum, \(\mathbf{W}^{\perp}\) ); formally,
\(\mathbf{W} = \mathbf{W}^{\parallel} \mathbf{W}^{\perp}\). If the
matrix has \(N\) samples and \(D\) parts, then we can uniquely identify
\(\mathbf{W}_{dn}\) as the \(d^{th}\) feature in sample \(n\). Any data
that we collect by sequencing is thought to contain only proportional
(Lovell et al. 2011) information and we can represent the corresponding
sequenced matrix as \(\mathbf{Y}\). Compositional data analysis makes
the strong assumption that
\(\mathbf{Y}_{n} \equiv \mathbf{Y}^{\parallel}_{n} = f(\mathbf{W}^{\parallel}_{n})\);
in other words that what is measured by sequencing is relative
information only, and that this measurement is a function of the
relative information in the underlying environment. Note that under this
assumption no corresponding estimate or information is available for the
scale of the sample \(\mathbf{Y}^{\perp}_{n}\).

One issue that was realized very early was that if the goal was to
compare one sample to another then the output from HTS needed to be
normalized in order to make the samples commensurate and to correct any
minor asymmetries in the data. A large number of normalizations were
developed. Rarefaction(Hughes and Hellmann 2005) was used in the
microbiome field, while transcriptome investigators developed and used
proportions and derivatives (reads per kilobase per million, and
transcripts per kilobase per million(Mortazavi et al. 2008; Wagner, Kin,
and Lynch 2012)), the relative log expression (RLE)(Anders and Huber
2010) and trimmed mean of M values (TMM)(Robinson and Oshlack 2010). The
centered log ratio (CLR)(Aitchison 1982) approach was proposed as a
unifying normalization(Fernandes et al. 2014). With the exception of
rarefaction, all of these normalizations are ratios with the major
differences between approaches being how the denominator is chosen and
whether the ratio is always log transformed or not.

Recently, Nixon et al. (2023) showed that that \emph{all} normalizations
make some identifying assumption about the scale of \(\mathbf{W}\) but
that the actual assumption being made was not always easy to interpret.
In essence, normalizations in widespread use assume that either one
sample can be chosen as a reference to which the others are scaled
(TMM), or that different sub-parts of each sample maintain a constant
scale across samples (RLE, CSS, CLR and derivatives).

The CLR, which can be written as
\(\mathbf{CLR}_{(1...d)n} = \mathbf{Y}_{(1...d)n} / G_n\), uses the
geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\) to scale the
data. Nixon et al. (2023) discovered that using the CLR was equivalent
to making a statement that \(G_n\) was related to the scale of the
underlying environment \(\mathbf{W}^{\perp}\). Interestingly, the
information in a system is a fundamental property and \(G_n\) is
strongly correlated with Shannon's entropy, indicating that information
about the complexity of \(\mathbf{W}\) is available post-sequencing
(Supplemental). The realization that all normalizations made assumptions
of scale led to the understanding that all normalizations introduced
latent bias into analyses because the scaling factor has been assumed up
to now to be a perfect point estimate of the scale of the underlying
system (Nixon et al. 2023).

With this new insight Nixon et al. (2023) devised methods to set up a
scale model that explicitly acknowledges and accounts for the
uncertainty of the scale. In the case of the CLR which uses the \(G_n\)
as the denominator, this entails adding uncertainty into \(G_n\) such
that the latent bias can be exposed, acknowledged and accounted for.
Additionally it was realized that the actual values used in the
denominator to calculate the log-ratio were irrelevant, but instead that
the ratios between the denominators was the key parameter. These two
insights allow for a full scale model of scale to be built and tested
for its effect on analysis.

The ALDEx2 R package (Fernandes et al. 2013) represents a general
purpose toolbox for Bayesian estimation of HTS datasets. ALDEx2 was
designed originally to convert the discrete point estimates in the count
matrix obtained into continuous posterior distributions through
Monte-Carlo sampling from the Dirichlet distribution. The posterior more
accurately represents variation on a per-sample basis(Gloor et al. 2016)
and this \(\mathbf{Y}^{\parallel}_{dn}\) can more closely approximate
\(\mathbf{W}^{\parallel}_{dn}\). Each Monte-Carlo replicate is
normalized using the CLR and used to calculate summary and test
statistics, and finally expected values and confidence intervals are
reported across the Monte-Carlo replicates. The CLR calculation can be
modified trivially to include a full scale model of the data by
incorporating uncertainty into \(G\) and so produce a posterior estimate
of the scale. The scale uncertainty can be included on a per-experiment,
per-group or even a per-sample basis. Thus, ALDEx2 can now be used to
generate a full posterior model of both \(\mathbf{W}^{\parallel}_{dn}\)
and \(\mathbf{W}^{\perp}_{dn}\) from the observed data.

@@

The first dataset is a highly replicated yeast transcriptome where one
condition is wild-type and the other has a snf-1 gene
knockout(Gierliński et al. 2015). Yeast deficient for snf-1 grow more
slowly and are sensitive to a variety of common agents that cause cell
stress (Yoshikawa et al. 2011). This dataset has been used to argue that
only a small number of replicates need to be used to identity
differentially abundant genes and that different tools should be used
for datasets with different sample sizes because the tools have
different intrinsic statistical power and Type 1 and Type 2 errors
(Schurch et al. 2016). This guidance runs counter to standard
statistical practice where power is intrinsically linked to sample size
(Halsey et al. 2015), yet the concept of sample-size independent power
is entrenched in all fields that use HTS as an experimental readout.
Increasing false precision with increasing sample size can be understood
as the result of unacknowledged bias brought about because of false
certainty as to the scale of the data. Thus, by adding scale uncertainty
we can be more confident that the analysis is not converging on a
precise but biased estimate of scale (Gustafson 2015).

We start by examining the dispersion or variance of the data as counts
and as the logarithm of the normalized counts as calculated by DESeq2
(log(RLE)) and by ALDEx2 (CLR). The actual counts of data derived from
sequencing are overdispersed with the mean value being less than the
variance (Robinson, McCarthy, and Smyth 2010) as seen in Panel A of
Figure 1. This relationship is why most tools model the counts with the
Negative Binomial distribution(Frazee et al. 2023), and use this
distribution as the basis for batch correction(Zhang, Parmigiani, and
Johnson 2020). However, the actual analysis of differential abundance is
performed on the logarithm of the normalized counts (Robinson, McCarthy,
and Smyth 2010; Love, Huber, and Anders 2014; Andrews and Hemberg 2019),
or on the clr values (Fernandes et al. 2013) both of which are
log-ratios. The mean-dispersion distribution of these log-ratio
transformed data is quite different as shown in panels B and C of Figure
1 and as noted elsewhere (Fernandes et al. 2013; Love, Huber, and Anders
2014).

@@

Here we can see that while the variance or dispersion always increases
with increasing raw read count, it actually decreases when measured on
the log-ratio transformed data (Fernandes et al. 2013; Love, Huber, and
Anders 2014) and reaches a minimum at some mid-point of the
distribution. This makes the counter-intuitive suggestion that genes
with moderate expression have more predictable expression than genes
with very high expression such as housekeeping genes. This is at odds
with the known biology of cells where single cell counting of
housekeeping transcripts shows that they are both highly expressed and
have little intrinsic variation (Taniguchi et al. 2010). Furthermore,
the actual amount of dispersion is very small and for many transcripts
is almost negligible. To show this point more clearly, the majority of
the transcripts in the lowest decile of dispersion indicated below the
dashed grey line are statistically significantly different (75\% with
DESeq2, 69\% with ALDEx2), suggesting that low dispersion estimates lead
to many false positives because of the underestimation of dispersion.
The work of Nixon et al. (2023) suggests that this is because the biased
scale estimates that are calculated. Thus, the actual variation of
highly expressed genes is not captured accurately by current approaches
to analysis and supports the idea that HTS have unacknowledged bias.

Using either DESeq2 or ALDEx2, we observe that a majority of transcripts
are statistically significantly different between groups even with a
Benjamini-Hochberg (Benjamini and Hochberg 1995) false discovery rate
(FDR) of 0.01; i.e.~4264 or 3790 of the 5891 transcripts are
statistically significant. Clearly, the necessary assumption of most
features being invariant is not justified. Applying the rule of thumb of
at least a \(2^{1.4}\) fold change (Schurch et al. 2016) reduces these
outputs to 193 for DESeq2 and to 186 for ALDEx2.

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Normalizations are not what you think; Explicit Scale Simulation in ALDEx2},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Normalizations are not what you think; Explicit Scale Simulation
in ALDEx2}
\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin D. Silverman$^{2,3,4}$}
\date{\today}

@@

$^1$ Department of Biochemistry, University of Western Ontario
$^2$ College of Information Sciences and Technology, Pennsylvania  State University
$^3$ Department of Medicine, Pennsylvania  State University
$^4$ Institute for Computational and Data Science, Pennsylvania State University

\begin{abstract}
In many high-throughput sequencing (HTS) studies, sample-to-sample
variation in sequencing depth is not driven by variation in the scale
(e.g., total size, microbial load, or total gene expression) of the
underlying biological systems being measured but rather is driven by
technical factors. Typically the technical variation is addressed using
some form of statistical normalization. Normalizations are data or
parameter transformations that remove unwanted technical variation in
the hopes of facilitating analyses sensitive to scale; e.g.,
differential abundance and differential expression analyses. Recently we
showed that any normalization makes implicit assumptions about the
unmeasured system scale and that errors in these assumptions can lead to
dramatic increases in false positive and false negative rates. Here we
describe updates to the ALDEx2 R package that mitigate these problems by
directly modeling uncertainty in the unmeasured system scale through the
use of a \textit{scale model}. Scale models generalize the idea of
normalizations and can be thought of as explicitly modeling potential
error in the chosen normalization. Beyond enhancing the robustness of
HTS analyses, the use of scale models within ALDEx2 enhances the
transparency and reproducibility of analyses by making implicit
normalizing assumptions an explicit part of the model building process.
\end{abstract}

@@

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to the flow cell. Each of these steps can be though
of as a sampling step: a pool of DNA that could be measured is sampled
and only that subsampled DNA is carried over to subsequent steps. Within
each sampling step, the connection between the size of the sampled DNA
pool and the scale (e.g., size, microbial load, or total gene
expression) of the measured biological system is degraded or lost.
Increasingly, researchers are turning to modified experimental protocols
(e.g., DNA spike-ins or cell counting) in an attempt to recover the lost
biological variation in scale; see for example (Vandeputte et al. 2017;
Props et al. 2017). However, these protocols often do not recover
meaningful biological variation but instead recover scale variation at
some intermediate step in the sample preparation protocol (e.g., at the
stage where the DNA spike-in was added). In short, the disconnect
between sample-to-sample variation in sequencing depth and biological
variation in scale remains an outstanding challenge.

Biological variation in scale can represent an important unmeasured
confounder in HTS analyses (Lovell et al. 2015). For example, cells
transformed by the cMyc oncogene have about 3 times the amount of mRNA
and about twice the rRNA content than non-transformed cells (Nie et al.
2012), and this dramatically skews transcriptome analysis (Lovén et al.
2012). In addition, wild-type and mutant strains of cell lines, yeast or
bacteria often have different growth rates, which would affect our
ability to identify truly differentially abundant genes (Yoshikawa et
al. 2011). As another example, the total bacterial load of the vaginal
microbiome differs by 1-2 orders of magnitude in absolute abundance
between the healthy and bacterial vaginosis states (Zozaya-Hinchliffe et
al. 2010), and the composition between these states is dramatically
different (Ravel et al. 2011; Hummelen et al. 2010). Thus, a full
description of these systems includes both relative change (composition)
and absolute abundance (scale), yet current methods access only the
compositional information.

To be more concrete, let \(\mathbf{Y}\) denote an \(D \times N\) matrix
of sequence counts with elements \(\mathbf{Y}_{dn}\) denoting the number
of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or
gene) in sample \(n\). Let \(\mathbf{W}\) also represent a
\(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing
the true (as opposed to measured) amount of class \(d\) in the
biological system from which sample \(n\) was obtained. Succinctly,
\(\mathbf{Y}\) denotes the observed HTS data whereas \(\mathbf{W}\)
denotes the true counts in the system under study. We can think of
\(\mathbf{W}\) as consisting of two parts, the scale
\(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
\(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
\(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
Note that with these definitions \(\mathbf{W}\) can be written as the
element-wise combination of scale and composition:
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
The technical variation in sequencing depth
(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
observed data \(\mathbf{Y}\) provides us with information about the
system composition \(\mathbf{W}^{\parallel}\) but little to no
information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
al.~2011).

Early on, researchers recognized the potential confounding effects of
the non-biological variation in sequencing depth and approached the
problem using various forms of \textit{normalization}. The idea of
normalization is to remove this variation thus making the samples
commensurate and to correct any minor asymmetries in the data. A large
number of normalizations were developed. Rarefaction (Hughes and
Hellmann 2005) was used in the microbiome field, while transcriptome
investigators developed and used proportions and derivatives (reads per
kilobase per million, and transcripts per kilobase per million)
(Mortazavi et al. 2008; Wagner, Kin, and Lynch 2012), the relative log
expression (RLE) (Anders and Huber 2010) and trimmed mean of M values
(TMM) (Robinson and Oshlack 2010). The Centered Log Ratio (CLR)
(Aitchison 1982) was proposed as a unifying normalization (Fernandes et
al. 2014). With the exception of rarefaction, all of these
normalizations are data transformations that can be stated as ratios of
the form \(\hat{\mathbf{Y}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\); the
major differences being how the denominator is chosen and whether the
ratio is subsequently always log-transformed or not.

Recently, Nixon et al. (2023) suggested that the challenge of
non-biological variation in sequencing depth be viewed as a problem of
partially-identified models. Through this lens, they showed that
\emph{all} normalizations make some assumption about the system scale
\(\mathbf{W}^{\perp}\) in order to identify the chosen model. Moreover,
these assumptions are often implicit, difficult to interpret, and
provided different outputs when applied to the same dataset (Dillies et
al. 2013; Weiss et al. 2017). Intuitively, normalizations in widespread
use assume that either one sample can be chosen as a reference to which
the others are scaled (TMM), or that different sub-parts of each sample
maintain a constant scale across samples (RLE, CLR and derivatives).
Further, Nixon et al. (2023) showed that the original ALDEx2 model made
a strict assumption about scale through the CLR normalization. They
showed that this assumption could be remedied since ALDEx2 permits scale
to be easily incorporated into its framework. In this paper, we discuss
these modifications and show how including scale uncertainty can greatly
improve modeling and lead to more robust and reproducible results.

\hypertarget{adding-scale-uncertainty-in-aldex2}{%
\subsection{Adding Scale Uncertainty in
ALDEx2}\label{adding-scale-uncertainty-in-aldex2}}

The ALDEx2 R package (Fernandes et al. 2013) is a general purpose
toolbox for Bayesian modeling of HTS data. For brevity and concreteness,
we discuss ALDEx2 in its simplest form as a tool for estimating
Log-Fold-Changes (LFC) (e.g., differential abundance or differential
expression analysis), but note that it can be used to fit more complex
linear models. At a high-level, ALDEx2 involves three steps: 1) a
Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given
observations \(\mathbf{Y}\); 2) a normalization is used to estimate
\(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the
\(\mathbf{W}\) estimates are used to estimate LFC as part of
differential abundance (or expression) analyses. For more details on
ALDEx2 see Fernandes et al. (2013).

By default, ALDEx2 uses the CLR normalization which can be written as
\(\log \mathbf{W}_{dn}=\log [\mathbf{W}^{\parallel}_{dn}/G_{n}]\) where
\(G_{n}\) denotes the geometric mean of the vector
\((W^{\parallel}_{1n}, \dots, W^{\parallel}_{Dn})\) (see Supplemental
Materials for how the CLR connects to Shannon entropy). Nixon et
al.~(2023) showed that in this context the CLR normalization equates to
an implicit assumption that \(\mathbf{W}^{\perp}_{n}=1/G_{n}\). They
showed that even slight errors in this assumption introduced bias into
LFC estimate that was not accounted for in estimates of uncertainty
(e.g., confidence intervals or credible sets). In fact, they showed that
the only way in which the ALDEx2 model, or any model that uses
normalizations, could ever be calibrated (e.g., control Type-I Error
rates) was if this assumption was exactly true. While this assumption
may be approximately correct in some instances it is surely never
exactly true. In support of this assertion it is known that false
discovery rates vary widely by analysis and normalization method
(Hawinkel et al. 2018; Li et al. 2022). As a result, Nixon et al.~(2023)
suggested that models should consider potential error in the assumptions
implied by normalizations.

Nixon et al.~(2023) proposed the concept of a \textit{scale model} as a
generalization of normalizations that account for potential error. They
showed how scale models can be incorporated into ALDEx2, turning the
ALDEx2 model into a specialized type of statistical model which they
called a \textit{Scale Simulation Random   Variable} (SSRV). Recall that
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).
As the HTS data only informs directly on
\(\mathbf{W}^{\parallel}_{dn}\), they proposed to augment analyses with
a model for \(\mathbf{W}^{\perp}_{n}\) (the scale model). Since the CLR
normalization makes the assumption \(\mathbf{W}^{\perp}_{n}=1/G_{n}\),
the CLR normalization can be generalized by considering probability
models for the scale \(\mathbf{W}^{\perp}_{n}\) that have mean
\(1/G_{n}\). For example, the following scale model generalizes the CLR:

\[\log \mathbf{W}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]

where \(\gamma\) is a tunable parameter that controls the degree of
uncertainty in the CLR assumption and \(x_{n}\) denotes a binary
condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\)
denotes control). Beyond theoretical studies, they demonstrated through
the analysis of simulated and real data the remarkable performance
improvements that could result from incorporating this type of scale
uncertainty into analyses. Subsequently, we have made those
modifications a permanent fixture of ALDEx2 which now represents the
first software package designed for SSRV-based inference.

Beyond concerns of fidelity and rigor, scale models also enhance the
reproducibility and transparency of HTS analyses. This occurs because
rather than using an implicit assumption (e.g.,
\(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale
models make this an explicit part of the model building process. While
it is beyond the scope of the present article, we note that there are
many ways of building scale models that enhance the interpretability of
the parameters and assumptions. Moreover, the scale model listed above,
which is now the default in ALDEx2, also has an appealing
interpretation. A detailed description of these points as well as the
mathematical underpinnings of the updated ALDEx2 R package is
forthcoming and will be made available as a companion publication. Here,
we simply treat the parameter \(\gamma\) in the default scale model as
an abstract term that represents uncertainty in the CLR assumption and
note that larger values of \(\gamma\) correspond to more uncertainty in
this assumption. We also show that the ratio of scales between
conditions can be used to build a full scale model when the conditions
have dramatically different underlying scales.

@@

\hypertarget{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}{%
\subsection{Adding scale uncertainty replaces the need for dual
significance
cutoffs.}\label{adding-scale-uncertainty-replaces-the-need-for-dual-significance-cutoffs.}}

Gierliński et al. (2015) conducted a highly replicated yeast
transcriptome experiment comparing a wild-type strain with a snf1 gene
knockout. This dataset has been used to argue that only a small number
of replicates are needed to identity differentially abundant genes and
makes the recommendation that some tools should be used in some
conditions because each tool has different intrinsic statistical power
and Type 1 and Type 2 errors (Schurch et al. 2016){]}. This guidance
runs counter to standard statistical practice where power is
intrinsically linked to sample size (Halsey et al. 2015), yet the
concept of sample-size independent power is entrenched in all fields
that use HTS as an experimental readout. Through the lens of this paper,
this is a hallmark of unacknowledged bias: a false certainty in the
scale of the data brings about an increasing false precision with
increasing sample sizes. Furthermore, the yeast transcriptome experiment
has an unacknowledged difference in scale between conditions: yeast
deficient for snf1 are smaller, grow more slowly and are sensitive to a
variety of common agents that cause cell stress (Yoshikawa et al.
2011){]}. Thus, by adding scale uncertainty, we can be more confident
that the results of the analysis reflect true biological differences and
not differences due to the assumptions made by the normalization method.

We start by examining the dispersion (variance) of the data as counts
and as the logarithm of the normalized counts when calculated by DESeq2
(log(RLE)) or by ALDEx2 (CLR). The raw data counts derived from
sequencing are overdispersed with the mean value being less than the
variance (Robinson and Oshlack 2010; Robinson, McCarthy, and Smyth 2010)
as seen in Panel A of Figure 1. However, the actual analysis of
differential abundance is performed on the logarithm of the normalized
counts (Robinson, McCarthy, and Smyth 2010; Love, Huber, and Anders
2014; Andrews and Hemberg 2019), which includes the CLR (Fernandes et
al. 2013) yet the mean-dispersion distribution of these log-ratio
transformed data is quite different as shown in panels B and C of Figure
1 and as noted elsewhere (Fernandes et al. 2013; Love, Huber, and Anders
2014).

@@

These relationships depend on what is being analyzed: the variance or
dispersion always increases with increasing raw (or normalzied) read
count but decreases when measured on the log-ratio transformed data
(Fernandes et al. 2013; Love, Huber, and Anders 2014), reaching a
minimum at some mid-point of the distribution. This makes the
counter-intuitive suggestion that genes with moderate expression have
more predictable expression than genes with very high expression such as
housekeeping genes. This is at odds with the known biology of cells
where single cell counting of housekeeping transcripts shows that they
are both highly expressed and have little intrinsic variation (Taniguchi
et al. 2010). Furthermore, the dispersion is exceeding small being, for
many transcripts, almost negligible. To show this point more clearly,
the majority of the transcripts in the lowest decile of dispersion
indicated below the dashed grey line are statistically significantly
different (75\% with DESeq2, 69\% with ALDEx2), suggesting that low
dispersion estimates lead to many false positives. From the perspective
of Nixon et al. (2023), it is reasonable to conclude that some of these
results may be due to the choice of normalization. Indeed, benchmarking
and comparison studies have repeatedly shown that the choice of
normalization plays a role in which results are returned as significant;
see for example (Maza et al. 2013), (Dillies et al. 2013) and (Weiss et
al. 2017).

Using either DESeq2 or ALDEx2, a majority of transcripts are
statistically significantly different between groups even with a
Benjamini-Hochberg (Benjamini and Hochberg 1995) false discovery rate
(FDR) of 0.01; i.e.~4264 or 3790 of the 5891 transcripts are
statistically significant. Clearly, a 70\% positive rate seems
biologically unjustified, and furthermore this breaks the necessary
assumption made by the normalization methods that most features must be
invariant. Finally, that 90 transcripts are identified by ALDEx2 and not
DESeq2, while DESeq2 identifies 564 transcripts that ALDEx2 does not
suggests that the choice of normalization plays a role in which results
are returned as significant.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Dillies:2013}{}}%
Dillies, Marie-Agnès, Andrea Rau, Julie Aubert, Christelle
Hennequet-Antier, Marine Jeanmougin, Nicolas Servant, Céline Keime, et
al. 2013. {``A Comprehensive Evaluation of Normalization Methods for
{Illumina} High-Throughput {RNA} Sequencing Data Analysis.''}
\emph{Brief Bioinform} 14 (6): 671--83.
\url{https://doi.org/10.1093/bib/bbs046}.
